---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis-CI Build Status](https://travis-ci.org/cboettig/EML.svg?branch=master)](https://travis-ci.org/cboettig/EML)


# EML 

*Note*: `EML` is work in progress. Please see the Issues tracker in this repository for details about current issues and development milestonds.

## Installation

`EML` has not yet been released.  Please install from GitHub after installing the `devtools` package (from CRAN):


```{r eval=FALSE}
devtools::install_github("cboettig/EML")
```


## Quickstart

Load the package and read in an EML file:

```{r}
library("EML")
f <- system.file("xsd/test/eml.xml", package = "EML")
eml <- read_eml(f)
```

This creates a native R object called `eml`.  Although this is an "S4 object" type, it uses a "show"
method which displays in an XML-like layout for convenience:

```{r}
eml
```

Validate EML against the official schema

```{r, message=TRUE}
# An EML document with no validation errors
eml_validate(eml)

# An EML document with validation errors
invalid_eml <- system.file("tests/testthat/", "example-eml-invalid.xml", package = "EML")

eml_validate(invalid_eml)
```


Write out as EML:


```{r}
write_eml(eml, "example.xml")
```


```{r include=FALSE, echo=FALSE}
unlink("example.xml")
```

## Manipulating EML objects

Eventually `EML` will provide constructor and extract methods to create and extract common metadata sections from convenient R stuctures (e.g. `data.frames` with unit metadata, common R classes like `Person`, `bibtype`).  For now, the only method to access and modify EML is to use the standard S4 subsetting and constructor methods.

Our current example does not have a publication date.  Let's add one:

```{r}
eml@dataset@ResourceGroup@pubDate <- new("pubDate", "2016")
```

Note that we use the constructor method `new()` to create an object.

### Working with repeating elements


## Developer notes

### Creating EML class definitions 

Class definitions (`classes.R`) and methods (`methods.R`) are created programmatically.  From the root of the package, run: `source("inst/create-package/create_package.R")`.  
