
In [Advanced writing of EML](), we construct richer EML files, including:

- Constructing more complete lists of authors, publishers and contact.
- Summarizing the geographic, temporal, and taxonomic coverage of the dataset
- Reading in pages of methods descriptions from a Word document
- Adding arbitrary additional metadata
- Indicating the canonical citation to the paper that should be acknowledged when the data is re-used.  
- Conversion between EML and other metadata formats, such as NCBII and ISO standards.

In so doing we will take a more modular approach that will allow us to build up our metadata from reusable components, while also providing a more fine-grained control over the resulting output fields and files.  




### Creating hf205.xml

In this example, we will use R to re-generate the EML metadata originally published by [Ellison _et al_ (2012)]() through the Harvard Forest Long Term Ecological Research Center.  <!-- The original metadata file was created in association with the publication in PNAS based on a Microsoft Word document template that Harvard Forest provides to the academic researchers.  Metadata from this template is then read off by hand and an EML file is generated using a combination of a commercial XML editing platform (Oxygen) for commonly used higher-level elements, and the Java platform `Morpho` provided by the EML development team for lower level attribute metadata. -->





#### Overview of the EML heirarchy

The EML metadata language 




We begin by reading in the CSV file providing the raw data that is to be annotated.  A copy of this file is distributed for demonstration purposes as part of the R package.  

```{r}
 dat <- read.csv(system.file("examples", "hf205-01-TPexp1.csv", package="reml"))
```

As in our [Basic tutorial](), essential metadata describes the columns and units used in the CSV file.  


```{r}
col.defs <- c("run.num" = "which run number (=block). Range: 1 - 6. (integer)",
              "year" = "year, 2012",
              "day" = "Julian day. Range: 170 - 209.",
              "hour.min" = "hour and minute of observation. Range 1 - 2400 (integer)",
              "i.flag" =  "is variable Real, Interpolated or Bad (character/factor)",
              "variable" = "what variable being measured in what treatment (character/factor).",
              "value.i" = "value of measured variable for run.num on year/day/hour.min.")
```



```{r}
unit.defs = list("which run number", 
                 "YYYY",
                 "DDD",
                 "hhmm", 
                 c(R = "real", I = "interpolated", B = "bad"),
                 c(control = "no prey added", 
                   low = "0.125 mg prey added ml-1 d-1",
                   med.low = "0,25 mg prey added ml-1 d-1",    # ERROR 
                   med.high = "0.5 mg prey added ml-1 d-1",
                   high = "1.0 mg prey added ml-1 d-1",
                   air.temp = "air temperature measured just above all plants (1 thermocouple)",
                   water.temp = "water temperature measured within each pitcher",
                   par = "photosynthetic active radiation (PAR) measured just above all plants (1 sensor)"),
                 c(control = "% dissolved oxygen",
                   low = "% dissolved oxygen",
                   med.low = "% dissolved oxygen",
                   med.high = "% dissolved oxygen",
                   high  = "% dissolved oxygen",
                   air.temp = "degrees C",
                   water.temp = "degrees C",
                   par = "micromoles m-1 s-1"))
```


We attach this essential metadata directly to the `data.frame` to make a `data.set` object (a data.frame with basic metadata).  


```{r}
dat <- data.set(dat, col.defs = col.defs, unit.defs = unit.defs)
```

Rather than write this out directly to an EML file with `eml_write` like we did in the basic example, we'll first assemble more metadata information.  

As a _metadata_ format, we do not encode the actual data file itself into the EML object.  Instead, the data is written out using some existing standard file format, which is merely _described by_ the EML file. The leading recommendation for exchanging and preserving tabular data is to use a plain text csv (comma separated values) file.  The function `eml_dataTable` will write out the data frame to a csv file, linking the filename to the metadata. We can optionally provide other information, such as description and name for the output file (if none is provided the unique id generated for this element will be used).  


```{r}
dataTable <- reml:::eml_dataTable(dat, 
                           description = "Metadata documentation for S1.csv", 
                           file = "S1.csv")
```

CSV file conventions can differ across platforms and continents. To remove any ambiguity, EML explicitly describes how the csv file is constructed (noting that it is actually comma delimited as the name implies and not tab delimited, and noting other details such as the number of rows, the character used to indicate and end of a row, and so forth).  Because R is writing the csv file itself after all, it already knows all of these details and we are able to automatically fill out this section of the metadata.  This information is summarized in the `dataTable@physical` element.  




One of the primary benefits in constructing EML in a scripted language such as R rather than in a GUI interface such as `morpho` is the ability to easily save and reuse components.  Here is an EML component that Harvard Forest will use a lot: it's address

```{r}
HF_address <- new("address", 
                  deliveryPoint = "324 North Main Street",
                  city = "Petersham",
                  administrativeArea = "MA",
                  postalCode = "01366",
                  country = "USA") 
```

We use the `new` function to construct an object of class `address`.  This R object has "slots" corresponding to each of the elements that makes up an address in EML.  We can define the values for these slots all at once as shown, or we can add and edit them after the object is created by using R's S4 subsetting operator, `@`.  For instance, we could change the street address with: `HF_address@deliveryPoint <- "some new address"`. (Some users might find the equivalent notation, `slot(HF_address, "deliveryPoint") <- "some new address"` to be more intuitive).  We need not specify any additional values when first calling `new("address")`, and we can use `slotNames(HF_address)` to see a list of possible fields.  

This element is now available for reuse as we create other metadata.  For instance, we will define the publisher of the data by providing a name and an address:

```{r}
publisher <- new("publisher", 
    organizationName = "Harvard Forest",
    address = HF_address)
```

Note that the `address` slot is not simply a text field, but is in fact an `address` S4 object we created above.  This object preserves the structure defined by the EML schema, which ensures that software programs can consistently interpret the values we write into the metadata file.  


We can reuse data formats native to R as well.  For instance, R comes with an inherent notion of a "person" as something with a given name, a family name, an email address, and so forth.  R recognizes a string written like this: 


```{r}
aaron <- as.person("Aaron Ellison <fakeaddress@email.com>")
```

can be interpreted as person's first and last name and email address.  The EML R package also recognizes this syntax, and can coerce this into the various 'person' or 'responsibleParty' roles defined in the EML schema.  Here we construct a contact node using coercion (the `as` function) from the `person` object

```{r}
contact <- as(aaron, "contact")
contact@address = HF_address
contact@organizationName = "Harvard Forest"
contact@phone = "000-000-0000"
```

We could have entered each slot manually using `new("contact")` as we did for the above elemnts. 

Multiple individuals can be listed as the creator (author) of data.  Here we use coercion (`as`) to turn strings into creator nodes, and then use R's concatenate operator, `c`, to combine them. 

```{r}
creator <- c(as("Aaron Ellison", "creator"), as("Nicholas Gotelli", "creator"))
```


Yet another way to create person objects is to use the helper function `eml_person`.  This function recongnizes the plain-text string conventions already used by R's person objects (or the person objects themselves, see ?person).   Here we add other researchers as contributors to the data.  The `eml_person` function automatically decides whether to return a `contact`, `creator` list, or `associatedParty`, based on the additional information provided (such as an email address in angle braces, contributor role in square brackets).  

```{r}
other_researchers <- eml_person("Benjamin Baiser [ctb]", "Jennifer Sirota [ctb]") # creates a 'ListOfassociatedParty' (roles != cre)
```

Again, we could just as well have created these nodes using any of the other mechanisms shown above.  


Some metadata is just a string: 

```{r}
pubDate <- "2012" 
title <- "Thresholds and Tipping Points in a Sarracenia Microecosystem at Harvard Forest since 2012"
```

When metadata fields have more complicated types, the package provides constructor functions to help build these elements succinctly.  For instance, while we could add our keyword sets manually using only `new`:

```{r eval=FALSE}
keys <-
  c(new("keywordSet", 
      keywordThesaurus = "LTER controlled vocabulary",
      keyword = c(new("keyword", keyword="bacteria"), new("keyword", keyword="carnivorous plants"), ...)
    ),
   new("keywordSet", 
      keywordThesaurus = "LTER core area",
      keyword = ...) 
```

and so forth, it is more compact to use the provided constructor function:

```{r}
keys <- eml_keyword(list("LTER controlled vocabulary" = c("bacteria", "carnivorous plants", "genetics", "thresholds"),
                                     "LTER core area" = c("populations", "inorganic nutrients", "disturbance"),
                                        "HFR default" = c("Harvard Forest", "HFR", "LTER", "USA")))
```

A similar situation arises with describing the coverage metadata element. 
```{r}
coverage <- eml_coverage(scientific_names = "Sarracenia purpurea", 
                         dates            = c('2012-06-01', '2013-12-31'),
                         geographic_description = "Harvard Forest Greenhouse, Tom Swamp Tract (Harvard Forest)", 
                         NSEWbox          = c( 42.55,  42.42, -72.1, -72.29, 160, 330))
```

In this case, the helper function supports only the more common types of coverage, such as listing taxonomic coverage at the species name rather than higher taxonomic level, or providing only calendar dates instead of, say, geological epochs.  For these alternative constructions, users will call the `new` constructors directly, see `?temporalCoverage`, `?taxonomicCoverage` and related functions for details.  


The remaining metadata provided in hf205.xml consists primarily of text blocks.  We might write small text blocks directly into our script:  

```{r}
abstract <- "The primary goal of this project is to determine
  experimentally the amount of lead time required to prevent a state
  change. To achieve this goal, we will (1) experimentally induce state
  changes in a natural aquatic ecosystem - the Sarracenia microecosystem;
  (2) use proteomic analysis to identify potential indicators of states
  and state changes; and (3) test whether we can forestall state changes
  by experimentally intervening in the system. This work uses state-of-the
  art molecular tools to identify early warning indicators in the field
  of aerobic to anaerobic state changes driven by nutrient enrichment
  in an aquatic ecosystem. The study tests two general hypotheses: (1)
  proteomic biomarkers can function as reliable indicators of impending
  state changes and may give early warning before increasing variances
  and statistical flickering of monitored variables; and (2) well-timed
  intervention based on proteomic biomarkers can avert future state changes
  in ecological systems."  
```

or read them in from some externa plain text file such as `abstract <- readLines("abstract.txt")`.  

```{r}
rights <- "This dataset is released to the public and may be freely
  downloaded. Please keep the designated Contact person informed of any
  plans to use the dataset. Consultation or collaboration with the original
  investigators is strongly encouraged. Publications and data products
  that make use of the dataset must include proper acknowledgement. For
  more information on LTER Network data access and use policies, please
  see: http://www.lternet.edu/data/netpolicy.html."
```


For larger blocks of text we might rather write these Word.  We can read them in from there, with the help of a few additional R packages.   

```{r}
library(RWordXML)
library(XML)
f2 <- wordDoc(system.file("examples", "methods.docx", package="reml"))
doc <- f2[[getDocument(f2)]]
txt <- xpathSApply(doc, "//w:t", xmlValue)
## FIXME add <title> <section> and <para> blocking back: 
method_description <- paste(txt, collapse = "\n\n") 
```

We can then construct a methods node containing this text as the description: 

```{r}
methods <- new("methods")
methods@methodsStep@description <- method_description
```


One of the simplest ways and most powerful ways to generate EML metadata is to resuse a section from an existing EML file directly.  

```{r}
additionalMetadata <- hf205@additionalMetadata # extracted from previous eml file
```




```{r}
dataset <- new("dataset", 
                title = title,
                creator = creator,
                contact = contact,
                pubDate = pubDate,
                associatedParty = other_researchers,
                keywordSet = keywords,
                coverage = coverage,
                methods = methods,
                dataTable = c(dataTable))

```

```{r}
eml <- new("eml", 
            dataset = dataset,
            additionalMetadata = additionalMetadata)
```

This `eml` object we have just created now has the complete R representation of the EML data.  This is the same class of object returned by `eml_read`, and consequently we can use all of the read methods provided to explore or manipulate it.  

Finally we are ready to write out our EML object to an XML file. 

```{r}
eml_write(eml, "hf205_from_reml.xml")
```

When we used the `eml_write()` function for the first time in the [Basic tutorial](), we simply passed a `data.set` object.  Internally the function made the remaining calls to `new("dataTable")`, `new("dataset")`, and `new("eml")` for us.  This time, the `eml_write` function can recognize that this is already an `eml` object, which it immediately converts to the XML representation and writes out as a file.  (If no file name is provided, the XML text will be returned to the console).  


As before, we can now validate our EML document to ensure all the information has been formatted correctly:

```{r}
eml_validate("hf205_from_reml.xml")
```

While it is difficult to create invalid EML using the constructor functions, it is not impossible so it never hurts to check.  As illustrated in the basic tutorial, we can now publish this file automatically to a central repository such as the KNB.   






