# One simple way to share your data


```{r, include=FALSE}
source("~/.knitr_defaults.R")
options(xtable.type = 'html')
```

Use `reml`.

You have your data as a `data.frame` in R, convient for your analyses and graphs.  How best to archive share that data in a way that will maximize its utility to your lab group, your future self?  What will most impress your NSF data management review, or improve the chances that others will use and cite your data set?  Good metadata.

White et al. outline "Nine simple ways to make it easier to (re)use your data".  Here we present one simple way to meet these criteria.  Before we worry about what the criteria are, let's just jump in and make our archival data shared:


For sake of argument, imagine a data.frame looking something like this:

```{r, results="asis"}
dat = data.frame(river=c("SAC", "SAC", "AM"),
                 spp = c("king", "king", "ccho"),
                 stg = c("smolt", "parr", "smolt"),
                 ct =  c(293L, 410L, 210L))

xtable::xtable(dat)
```


In case our column header abbreviations are not obvious to others (or our future selves), we take a moment to define them: 


```{r}
col_metadata = c(river = "http://dbpedia.org/ontology/River",
                 spp = "http://dbpedia.org/ontology/Species",
                 stg = "Life history stage",
                 ct = "count")
```


Define the units used in each column, or the abbreviations we use in factors.  
 
```{r}
unit_metadata = 
  list(river = c(SAC = "The Sacramento River", AM = "The American River"),
       spp = c(king = "King Salmon", ccho = "Coho Salmon"),
       stg = c(parr = "third life stage", smolt = "fourth life stage"),
       ct = "number")
```


Specify the basic file metadata of author/contact, title, and a description, and choose where to publish:


```{r}
eml_write(dat, col_metadata, unit_metadata, 
          .contact="Carl Boettiger <cboettig@gmail.com>",  
          .title="A test EML file", 
          file_description="Test data, only intended for testing",
          destination="figshare")
```


We can provide much richer annotation than this example includes.  We can also store information such as the contact name and address in a configuration option to avoid having to provide it repeatedly.  Let's see how this example does on the Nine simple criteria. 


## 1. Share your data

Sharing is as easy as `publish_reml("my_eml_data.xml")`

The data is publically available at the DOI returned. At this DOI we find two files: the raw data in `.csv` format, and the EML data file providing machine readable metadata.  We'll unpack all the information provided in this metadata as we go along.  Many other options for sharing are possible, including private sharing and sharing through a variety of data publication platforms.  



## 2. Provide metadata

We provide `n` lines of metadata in the EML file, one of the "metadata standards" mentioned by White et al.  This includes basic information like the title, author(s), and contact address.  

> 1) the what, when, where, and how of data collection, 

Methods sections in the file provide answers to these questions.  Though our minimal example does not describe the collection itself, note that we nonetheless have generated a `methods` node, which describes the software that created the metadata file (after all, this is a step in the analysis).  

> 2) how to find and access the data, 

Our example returns a DOI and access URLs for the files in the data set: 

Rich access options are possible depending on the publication outlet, including specifying what users or groups have access to the data.  

> 3) suggestions on the suitability of the data for answering specific questions, 

We can see at a glance what taxa are covered by the data.  More detailed description can be provided in an abstract.

> 4) warnings about known problems or inconsistencies in the data, and 
> 5) information to check that the data are properly imported, such as the number of rows and columns in the dataset and the total sum of numerical columns

Checks have automatically been performed in generating the data, and are performed in reading in the data into R.  


## 3. Provide unprocsessed data

The second file provides the unprocessed data as a `.csv` file.  Because `csv` file formats vary


## 4. Use standard formats

Of the "nine ways", this one places the greatest burden on the creator and user.  Rather than coercing our data into prescribed formats, EML lets us simply describe whatever format we are using. 

- Standard file formats

Our example generate standard .csv file format, and declares this explicitly so that users can be sure what character is used to seperate values, rows, etc:

```xml
      <physical>
        <objectName>river_spp_stg_ct.csv</objectName>
        <dataFormat>
          <textFormat>
            <numHeaderLines>1</numHeaderLines>
            <recordDelimiter>\r\n</recordDelimiter>
            <attributeOrientation>column</attributeOrientation>
            <simpleDelimited>
              <fieldDelimiter>,</fieldDelimiter>
            </simpleDelimited>
          </textFormat>
        </dataFormat>
      </physical>
```

- Standard table formats
- Standard cell formats 

Rather than insist everyone use the ISO8601 format for dates, the user can declare whatever date format they prefer.  

## 5. Good null values

R has good null values in its use of `data.frames`, so we already have these.  Better yet, we write out our choice of null value in the metadata.  Reading in the data, we can explicitly pass this to `read.csv` rather than just assume the null value is appropriate.  This is handled automatically by `eml_read`.  

## 6. Make it easy to combine 

Context nodes serve the pupose described in this section.   
Machine-readable metadata, 

### 7. Perform basic quality control

We have quality control on the metadata as well as the data.  While user mistakes are still possible, automated validation of the metadata provided and 


## 8. Use an established repository

`reml` can automatically push our data to a variety of established repositories, as illustrated in this example. 

## 9. Licenses

We see by default `reml` has provided us with the liberal CC0 license for the data, consistent with major repositories such as Dryad.  Other options can be specified if needed.  






Ethan White et al have recently published an important paper on "Nine simple ways to share your data", which has garnered considerable attention even as a preprint [^1]




[^1]: Disclaimer: I reviewed this paper, Karthik Ram served as handling editor.  
